cardinalR: Generating Interesting High-Dimensional Data Structures

### 2025-12-13 paper-cardinalR.Rmd
Fix range notation in polynomial functions

The `gen_longlinear(n, p)` function generates a high-dimensional dataset representing a long linear structure with noise. Each variable is constructed by first creating a sequence of integers from $0$ to $n-1$, adding Gaussian noise with standard deviation $0.03n$, and then applying a random scaling and shift. Specifically, for the $j$-th dimension, a scale factor $\text{scale}_j \sim U(-10, 10)$ determines the orientation and stretching of the line, while a shift factor $\text{shift}_j \sim U(-300, 300)$ is added inside the multiplication to offset the line. This results in each variable being of the form $X_j = \text{scale}_j \cdot \big((0,1,\dots,n-1) + \epsilon + \text{shift}_j\big)$, where $\epsilon \sim N(0, (0.03n)^2)$ is independent Gaussian noise.

@@

The first is the quadratic curve of $n$ points in two dimensions. This is generated using `gen_quadratic(n, range)`. The independent variable is defined as $X_1 \sim U(\text{range}[1], \text{range}[2])$, and the response is generated as $X_2 = X_1 - X_1^2 + \varepsilon$, where $\varepsilon \sim U(0, 0.5)$. This produces a smooth parabolic arc opening downward, with vertical jitter introduced by the noise term.

@@

The second is the cubic curve of $n$ points in two dimensions. This is generated using `gen_cubic(n, range)`. The independent variable is defined as $X_1 \sim U(\text{range}[1], \text{range}[2])$, and a raw polynomial basis of degree $3$ is applied to construct $X_2 = X_1 + X_1^2 - X_1^3 + \varepsilon_2$, where $\varepsilon_2 \sim U(0, 0.5)$. This produces a more complex curvilinear structure than the quadratic case, with both upward and downward turning points.

>>>>

The `gen_longlinear(n, p)` function generates a high-dimensional dataset representing a single noisy linear trajectory. Let $t_i = i - 1, \quad i = 1, \dots, n$, denote a common latent index shared across all dimensions. For each dimension $j = 1, \dots, p)$, independent scale and shift parameters are sampled as $a_j \sim U(-10, 10), \qquad b_j \sim U(-300, 300)$. Gaussian noise $\varepsilon_{ij} \sim N(0, (0.03n)^2)$ is added independently across observations and dimensions. The observed variables are then defined as $X_{ij} = a_j \bigl(t_i + b_j + \varepsilon_{ij}\bigr), \quad i = 1, \dots, n$.cThis construction yields a single elongated linear structure embedded in $p\text{-}D$, with each dimension exhibiting a different orientation, scale, and offset.

@@

The first is the quadratic curve of $n$ points in two dimensions. This is generated using `gen_quadratic(n, range)`. Let $range = [a, b]$. The independent variable is defined as $X_1 \sim U(a, b)$, and the response is generated as $X_2 = X_1 - X_1^2 + \varepsilon$, where $\varepsilon \sim U(0, 0.5)$. This produces a smooth parabolic arc opening downward, with vertical jitter introduced by the noise term.

@@

The second is the cubic curve of $n$ points in two dimensions. This is generated using `gen_cubic(n, range)`. Let $range = [a, b]$. The independent variable is defined as $X_1 \sim U(a, b)$, and a raw polynomial basis of degree $3$ is applied to construct $X_2 = X_1 + X_1^2 - X_1^3 + \varepsilon_2$, where $\varepsilon_2 \sim U(0, 0.5)$. This produces a more complex curvilinear structure than the quadratic case, with both upward and downward turning points.

### 2025-12-13 paper-cardinalR.tex
Fix range notation in polynomial functions

The \texttt{gen\_longlinear(n,\ p)} function generates a high-dimensional dataset representing a long linear structure with noise. Each variable is constructed by first creating a sequence of integers from \(0\) to \(n-1\), adding Gaussian noise with standard deviation \(0.03n\), and then applying a random scaling and shift. Specifically, for the \(j\)-th dimension, a scale factor \(\text{scale}_j \sim U(-10, 10)\) determines the orientation and stretching of the line, while a shift factor \(\text{shift}_j \sim U(-300, 300)\) is added inside the multiplication to offset the line. This results in each variable being of the form \(X_j = \text{scale}_j \cdot \big((0,1,\dots,n-1) + \epsilon + \text{shift}_j\big)\), where \(\epsilon \sim N(0, (0.03n)^2)\) is independent Gaussian noise.

@@

The first is the quadratic curve of \(n\) points in two dimensions. This is generated using \texttt{gen\_quadratic(n,\ range)}. The independent variable is defined as \(X_1 \sim U(\text{range}[1], \text{range}[2])\), and the response is generated as \(X_2 = X_1 - X_1^2 + \varepsilon\), where \(\varepsilon \sim U(0, 0.5)\). This produces a smooth parabolic arc opening downward, with vertical jitter introduced by the noise term.

The second is the cubic curve of \(n\) points in two dimensions. This is generated using \texttt{gen\_cubic(n,\ range)}. The independent variable is defined as \(X_1 \sim U(\text{range}[1], \text{range}[2])\), and a raw polynomial basis of degree \(3\) is applied to construct \(X_2 = X_1 + X_1^2 - X_1^3 + \varepsilon_2\), where \(\varepsilon_2 \sim U(0, 0.5)\). This produces a more complex curvilinear structure than the quadratic case, with both upward and downward turning points.

>>>>

The \texttt{gen\_longlinear(n,\ p)} function generates a high-dimensional dataset representing a single noisy linear trajectory. Let \(t_i = i - 1, \quad i = 1, \dots, n\), denote a common latent index shared across all dimensions. For each dimension \(j = 1, \dots, p)\), independent scale and shift parameters are sampled as \(a_j \sim U(-10, 10), \qquad b_j \sim U(-300, 300)\). Gaussian noise \(\varepsilon_{ij} \sim N(0, (0.03n)^2)\) is added independently across observations and dimensions. The observed variables are then defined as \(X_{ij} = a_j \bigl(t_i + b_j + \varepsilon_{ij}\bigr), \quad i = 1, \dots, n\).cThis construction yields a single elongated linear structure embedded in \(p\text{-}D\), with each dimension exhibiting a different orientation, scale, and offset.

@@

The first is the quadratic curve of \(n\) points in two dimensions. This is generated using \texttt{gen\_quadratic(n,\ range)}. Let \(range = [a, b]\). The independent variable is defined as \(X_1 \sim U(a, b)\), and the response is generated as \(X_2 = X_1 - X_1^2 + \varepsilon\), where \(\varepsilon \sim U(0, 0.5)\). This produces a smooth parabolic arc opening downward, with vertical jitter introduced by the noise term.

The second is the cubic curve of \(n\) points in two dimensions. This is generated using \texttt{gen\_cubic(n,\ range)}. Let \(range = [a, b]\). The independent variable is defined as \(X_1 \sim U(a, b)\), and a raw polynomial basis of degree \(3\) is applied to construct \(X_2 = X_1 + X_1^2 - X_1^3 + \varepsilon_2\), where \(\varepsilon_2 \sim U(0, 0.5)\). This produces a more complex curvilinear structure than the quadratic case, with both upward and downward turning points.
